---
title: 'STAT 443: Homework 4'
author: 'Aronn Grant Laurel (21232475)'
date: "3 April, 2025"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#if you do not have the package, type install.packages("name_of_the_package")
library(knitr)
library(tseries)
library(ggplot2)
```

### Question 1
(a & b)


(c)

```{r, echo=TRUE}
# Calculate the spectral density

omega <- seq(-pi, pi, length.out = 500)
f_vals <- spectral_density(omega)

spectral_density <- function(omega) {
  (1 / (2 * pi)) * (1 / (1.05 - 0.16 * cos(omega) - 0.4 * cos(2 * omega)))
}

# Create a data frame for plotting
data <- data.frame(omega = omega, f_X = f_vals)

# Plot using ggplot2
ggplot(data, aes(x = omega, y = f_X)) +
  geom_line() +
  labs(title = "Spectral Density : AR(2) Process",
       x = "Frequency",
       y = "f(w)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 14))
```
We can the spectral density being dominated by low frequencies, with 
the spectral density is at highest at frequency = 0 as it's peak.
As we move further away from frequency 0, we observe
a dip / decreasing spectral density and another two smaller peaks
at approximately frequencies -3 and 3. Overall, this suggests that
our AR(2) process has long term dependencies.


### Question 2

(a) 

```{r, echo=TRUE}
# this is where your R code goes
```


(b)

```{r, echo=TRUE}
# this is where your R code goes
```
