---
title: 'STAT 443: Homework 3'
author: 'Aronn Grant Laurel (21232475)'
date: "14 March, 2025"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#if you do not have the package, type install.packages("name_of_the_package")
library(knitr)
library(tseries)
library(forecast)
```

### Question 1
```{r, echo=TRUE, fig.height = 7}
# this is where your R code goes

data <- read.csv("employee_wages_total_industry.csv")

data.ts <- ts(data$employee_wages,
              start = c(1997, 1),
              frequency = 12)

# January 1997 to December 2018 as the training set for model fitting
data.ts.train <- window(data.ts,
                        start = c(1997, 1),
                        end = c(2018, 12),
                        frequency = 12)
  
# from January 2019 to December 2019 as the test set
data.ts.test <- window(data.ts,
                       start = c(2019, 1),
                       end = c(2019, 12),
                       frequency = 12)

# LOESS Smoothing
data.ts.ls <- stl(data.ts.train, s.window = "periodic")

error.ls <- data.ts.ls$time.series[, "remainder"]
plot(error.ls)

# Model fitting for Remainder Term
error.time.ls <- seq_along(error.ls)
error.ls.df <- data.frame("Remainder" = error.ls,
                          "Time" = error.time.ls)

acf(as.vector(error.ls),
    lag.max = 50,
    main = "Correlogram of Error Term")

pacf(as.vector(error.ls),
     lag.max = 50,
     main = "PACF of Error Term")

```

From the time plot, we can observe an oscilation at value
0 while there are no obvious trend / pattern.

Our acf plot shows no consistency with a white noise 
process which suggests there is still some seasonal 
pattern in the remainder term / Process remembers itself.

Our PACF plot shows no consistency with a more 'damped'
wave / oscilation.

We observe a series with a cyclical component.


### Question 2
```{r, echo=TRUE}
# this is where your R code goes

```


### Question 3
```{r, echo=TRUE}
# this is where your R code goes
```


### Question 4
```{r, echo=TRUE}
# this is where your R code goes
```
