---
title: 'STAT 443: Lab 11'
author: 'Aronn Grant Laurel (21232475)'
date: "28 March, 2025"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#if you do not have the package, type install.packages("name_of_the_package")
library(knitr)
library(tseries)
```

### Question 1

(a) 

```{r, echo=TRUE}
set.seed(123)

wn <- arima.sim(model = list(order = c(0, 0, 0)), n = 100, sd = 2) 

# Before Smoothing
spec.pgram(wn, log = "no", main = " Periodogram")
abline(h = 1, col = 2, lty = 1, lwd = 2)

# After Smoothing (From Lec 6 P32)
spec.pgram(wn, spans = c(5), log = "no", main = "Smoothed Periodogram (spans = c(5))")

spec.pgram(wn, spans = c(15), log = "no", main = "Smoothed Periodogram (spans = c(15))")

spec.pgram(wn, spans = c(7, 5), log = "no", main = "Doubly Smoothed Periodogram (spans = c(7,5))")

```
For the first periodogram, we observe a more 'noisy' graph.
As we increase the length (m) of our smoothing filter from 
5 to 15, we observe an increase in smoothing where the 
noise reduces.
By applying the two step smoothing, we observe additional
smoothing effects on several peaks. Although Larger spans 
could provide a more percise estimate of the spectral 
density, but we must consider for over smoothing.


(b)

```{r, echo=TRUE, fig.height = 7}
set.seed(123)

wn_1000 <- arima.sim(model = list(order = c(0, 0, 0)), n = 1000, sd = 2)

# Before Smoothing / Raw Periodogram
spec.pgram(wn_1000, log = "no", main = "Raw Periodogram (n = 1000)")

# After Smoothing Periodogram
spec.pgram(wn_1000, spans = c(5), log = "no", main = "Smoothed Periodogram (spans = c(5))")

spec.pgram(wn_1000, spans = c(15), log = "no", main = "Smoothed Periodogram (spans = c(15))")

spec.pgram(wn_1000, spans = c(7,5), log = "no", main = "Smoothed Periodogram (spans = c(7,5))")


```
For our first periodogram before smoothing, we observe a 
'noisy' fluctuating graph because of noise. And as we 
increase our span size, we see our periodogram smoothing 
out more with less variability for span c(15). We also see 
a doubly smoothing which gives a relatively 'smoother' look
in the graph. Overall, we observe a smoothed periodogram 
with better approximation for our spectral density 
function by removing any extreme fluctuations.



(c)
### Setting logs = "yes" from question (b)
```{r, echo=TRUE, fig.height = 7}
set.seed(123)

wn_1000 <- arima.sim(model = list(order = c(0, 0, 0)), n = 1000, sd = 2)

# Before Smoothing / Raw Periodogram
spec.pgram(wn_1000, log = "no", main = "Raw Periodogram (n = 1000)")

# After Smoothing Periodogram
spec.pgram(wn_1000, spans = c(5), log = "yes", main = "Logged Smoothed Periodogram (spans = c(5))")

spec.pgram(wn_1000, spans = c(15), log = "yes", main = "Logged Smoothed Periodogram (spans = c(15))")

spec.pgram(wn_1000, spans = c(7,5), log = "yes", main = "Logged Smoothed Periodogram (spans = c(7,5))")
```
While our periodogram without log and smoothing will 
maintain the same, our logged and smoothed periodogram 
shows more compressed higher values with out spectrum range
decreasing. As we increase the smoothing, we observe less 
fluctuations and smoother curves. 
Our doubly logged and smoothed periodogram shows a 
relatively smoother and similar spectrum range.

I would believe span c(7, 5) be the 'best' choice in criteria
of consistent horizontal levels and relatively fewer dips and 
rises, meaning that it has a fair variance while also reducing 
noise well. 

(d)
From my findings in c), Taking the log helps reduce error by 
'compressing' the higher values. Hence, enabling the log 
transformation reduced the variance in the spectrum.


### Question 2
(a)
```{r, echo=TRUE, fig.height = 7}
set.seed(123)

ma_model <- arima.sim(n = 100, model = list(ma = -0.9), sd = 2)

# Raw periodogram
spec.pgram(ma_model, log = "no",main = "Periodogram of 
           AR(1) Process")

# Smoothed Periodogram
spec.pgram(ma_model, log = "no", spans = c(5), main =
             "Smoothed Periodogram (span = 5)")

spec.pgram(ma_model, log = "no", spans = c(15), main =
             "Smoothed Periodogram (span = 15)")

spec.pgram(ma_model, log = "no", spans = c(7,5), main =
             "Smoothed Periodogram (span = 7,5)")


```
As we increase the span 5, 7, and 15, we observe a lot smoother with 
less fluctuations (rise & dips) with 15 having the least rise and 
dips. Overall, we can observe a notice-able increase trend.
(b)
```{r, echo=TRUE, fig.height = 7}
set.seed(123)

true_spectrum <- function(omega) {
  (7.24 / pi) * ((1 - 1.8 * cos(omega)) / 1.81)
}

omega_vals <- seq(0, pi, length.out = 500)
theoretical_vals <- true_spectrum(2 * pi * omega_vals)

spec.pgram(ma_model, log = "no", main = "Raw Periodogram (n = 100)")
lines(omega_vals, theoretical_vals, col = "red", lwd = 2)

spec.pgram(ma_model, spans = c(5), log = "no", main = "Spans = c(5)")
lines(omega_vals, theoretical_vals, col = "red", lwd = 2)

spec.pgram(ma_model, spans = c(15), log = "no", main = "Spans = c(15)")
lines(omega_vals, theoretical_vals, col = "red", lwd = 2)

spec.pgram(ma_model, spans = c(7, 5), log = "no", main = "Spans = c(7,5)")
lines(omega_vals, theoretical_vals, col = "red", lwd = 2)

```

(c)
```{r, echo=TRUE, fig.height = 7}
set.seed(123)

ma_model2 <- arima.sim(n = 1000, model = list(ma = -0.9), sd = 2)

# Raw periodogram
spec.pgram(ma_model2, log = "no",main = "Periodogram of 
           AR(1) Process")

# Smoothed Periodogram
spec.pgram(ma_model2, log = "no", spans = c(5), main =
             "Smoothed Periodogram (span = 5)")

spec.pgram(ma_model2, log = "no", spans = c(15), main =
             "Smoothed Periodogram (span = 15)")

spec.pgram(ma_model2, log = "no", spans = c(7,5), main =
             "Smoothed Periodogram (span = 7,5)")

```
As we increase our smoothing span from 5 to 15, we start to observe 
a reduction in noise where it becomes much more smoother with less
fluctuations (rise & dips). We can see the transition from span 5 to 
15 where more high-frequency noises are reduced. 

(d)
```{r, echo=TRUE, fig.height = 7}
set.seed(123)

omega_vals <- seq(0, pi, length.out = 500)
theoretical_vals <- true_spectrum(2 * pi * omega_vals)

spec.pgram(ma_model2, log = "no", main = "Raw Periodogram (n = 100)")
lines(omega_vals, theoretical_vals, col = "red", lwd = 2)

spec.pgram(ma_model2, spans = c(5), log = "no", main = "Spans = c(5)")
lines(omega_vals, theoretical_vals, col = "red", lwd = 2)

spec.pgram(ma_model2, spans = c(15), log = "no", main = "Spans = c(15)")
lines(omega_vals, theoretical_vals, col = "red", lwd = 2)

spec.pgram(ma_model2, spans = c(7, 5), log = "no", main = "Spans = c(7,5)")
lines(omega_vals, theoretical_vals, col = "red", lwd = 2)

```
Since we used n = 1000, we expect the periodogram to be closer to the 
true spectral density than when n = 100. Observing our graphs, while 
increasing the span of the periodogram reduces variability for both 
n's, increasing n improves our periodogram's ability to capture the 
spectral density better.