---
title: 'STAT 443: Lab 11'
author: 'Aronn Grant Laurel (21232475)'
date: "28 March, 2025"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#if you do not have the package, type install.packages("name_of_the_package")
library(knitr)
library(tseries)
```

### Question 1

(a) 

```{r, echo=TRUE}
set.seed(1) 
wn <- arima.sim(model = list(order = c(0, 0, 0)), n = 100, sd = 2) 

# Before Smoothing
spec.pgram(wn, log = "no", main = " Periodogram")

# After Smoothing 
spec.pgram(wn, spans = c(5), log = "no", main = "Smoothed Periodogram (spans = c(5))")

spec.pgram(wn, spans = c(15), log = "no", main = "Smoothed Periodogram (spans = c(15))")

spec.pgram(wn, spans = c(7, 5), log = "no", main = "Smoothed Periodogram (spans = c(7,5))")


```
For the first periodogram, we observe a more 'noisy' graph.
As we increase the length (m) of our smoothing filter from 
5 to 15, we observe an increase in smoothing where the 
noise reduces.
By applying the two step smoothing, we observe additional
smoothing effects on several peaks. Although Larger spans 
could provide a more percise estimate of the spectral 
density, but we must consider for over smoothing.


(b)

```{r, echo=TRUE}
# this is where your R code goes
```


(c)

```{r, echo=TRUE}
# this is where your R code goes
```

