---
title: 'STAT 443: Lab 5'
author: 'Aronn Grant Laurel (21232475)'
date: "7 February, 2025"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#if you do not have the package, type install.packages("name_of_the_package")
library(knitr)
library(tseries)
library(zoo)
```

### Question 1
The process in equation 1 is an AR(3) Process with an order p = 

p(3) = alpha^(k) for k = 1, 2, 3
alpha^(1) = 0.8
alpha^(2) = -1/3
alpha^(3) = 0.6/sqrt(3)

```{r, echo = TRUE}
# Process AR(3) order in R studio
alpha1 <- 0.8
alpha2 <- -1/3
alpha3 <- 0.6 / sqrt(3)

sigma <- 0.8
```

alpha^(1) = 0.8
alpha^(2) = -1/3
alpha^(3) = 0.6 / sqrt(3)


### Question 2
Explain how to recognize this process based on an observed time series and 
how to determine its order.

We need to examine for stationarity by its constant mean and invertibility 
along with it's invertibility. We can also analyze through the ACF plot, 
where a AR(3) process should show a gradual / smooth decaying pattern.
In addition, we need to plot its PACF when it cuts off at lag 3 and ensure that
there are no autocorrelation.



### Question 3
Use the command set.seed(23456) to set the random seed for reproducibility and then use function arima.sim() to generate 1500 observations from the model in (1). Plot the simulated time series.

```{r, echo=TRUE}
set.seed(23456)

q3 <- arima.sim(model = list(ar = c(alpha1, alpha2, alpha3)), n = 1500, sd = sigma)
plot(q3, xlab = "Value", ylab = "Time")

```

### Question 4
```{r, echo=TRUE}
q3_ts <- ts(q3)
acf(q3_ts, main = "ACF")
```

The sample ACF plot shows a gradual decay pattern starting at lag 0 with some
osciliation after lag 16. After lag 14, we start to see insignificant results
within the boundary line. I believe that our results align with our AR(3) model
because we see alternating signs which would cause osciliation and the decaying
pattern is also exhibited in an AR model.

### Question 5
```{r, echo=TRUE}
pacf(q3_ts, main = "PACF")
```




### Question 6
```{r, echo=TRUE}

# Simulate AR(3) process
Z <- rnorm(1500, mean = 0, sd = 0.8)  # White noise
X <- numeric(1500)  # Initialize time series

# Generate time series data
for (t in 4:n) {
  X[t] <- alpha1 * X[t-1] + alpha2 * X[t-2] + alpha3 * X[t-3] + Z[t]
}

# Fit AR(3) model with different estimation methods
model_css_ml <- arima(X, order = c(3, 0, 0), method = "CSS-ML")
model_ml <- arima(X, order = c(3, 0, 0), method = "ML")
model_css <- arima(X, order = c(3, 0, 0), method = "CSS")

# Extract parameter estimates
results <- data.frame(
  Estimation = c("True Values", "CSS-ML", "ML", "CSS"),
  alpha1 = c(alpha1, coef(model_css_ml)[1], coef(model_ml)[1], coef(model_css)[1]),
  alpha2 = c(alpha2, coef(model_css_ml)[2], coef(model_ml)[2], coef(model_css)[2]),
  alpha3 = c(alpha3, coef(model_css_ml)[3], coef(model_ml)[3], coef(model_css)[3]),
  Variance = c(sigma, model_css_ml$sigma, model_ml$sigma, model_css$sigma)
)

# Round to 3 decimal places
results[, -1] <- round(results[, -1], 3)

# Present results in a table using kable
kable(results, caption = "Comparison of AR(3) Parameter Estimates Using Different Estimation Methods")

```



