---
title: 'STAT 443: Lab 8'
author: 'Aronn Grant Laurel (21232475)'
date: "8 March, 2022"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#if you do not have the package, type install.packages("name_of_the_package")
library(knitr)
library(tseries)
library(zoo)
library(forecast)
```

### Question 1

(a) 

```{r, echo=TRUE}
# this is where your R code goes
souvenir <- scan("souvenir.txt")

souvenir_ts <- ts(souvenir,
                  start = c(1987, 1),
                  end = c(1993, 12),
                  frequency = 12)

train_data <- window(souvenir_ts, 
                     start = c(1987, 1),
                     end = c(1993, 1))

test_data <- window(souvenir_ts,
                    start=c(1993, 2),
                    end=c(1993, 12))



plot(souvenir_ts, 
     main = "Monthly Sales of Souvenir Shop", 
     ylab = "Sales (A$)",
     xlab = "Time")

acf(souvenir_ts, main = "Sample ACF of Souvenir Sales")


```
There is a seasonal effect as we can see consistent periodic spikes in the monthly sales. We also see that
the size of the spikes increase over time and is not consistently the same size which suggests a multiplicative
model.


(b)

```{r, echo=TRUE}
# this is where your R code goes
holt_model <- HoltWinters(train_data,
                          seasonal = "multiplicative")
holt_model

plot(holt_model,
     main = "Holt-Winters Filtering: Souvenir Sales",
     ylab = "Sales (A$)",
     xlab = "Time")
```

$\alpha$ = 0.3746875 
$\beta$ = 0.04573451
$\gamma$ = 0.4522636

### Question 3
```{r, echo=TRUE, fig.height = 7}
# this is where your R code goes
plot(fitted(holt_model), main = "Fitted HoltWinters Model")
```

### Question 4
```{r, echo=TRUE, fig.height = 7}
# this is where your R code goes
predictions <- predict(holt_model,
                       n.ahead = length(test_data),
                       prediction.interval = TRUE)

plot(test_data,
     col = "black",
     type="b",
     pch=19,
     main = "Monthly Sales: Actual vs Predicted (1993)",
     ylab = "A$", xlab = "Time")


lines(predictions[, "fit"],
      type="b",
      pch=19,
      col = "blue",
      lty = 1)

lines(predictions[, "lwr"],
      type = "l",
      col = "red",
      lty = 2)

lines(predictions[, "upr"],
      type = "l",
      col = "red",
      lty = 2)

legend("topleft", 
       legend = c("Actual", "Predicted", "95% Prediction
                  Interval"),
       col = c("black", "blue", "red"),
       lty = c(1, 1, 2))

```
model follows the data fairly well and the real value is
usually within the prediction interval except for the
first data point (February 1993). Some data points were on the
edge of the prediction intervals where the
prediction wasn’t as accurate: March 1993, May 1993.

### Question 5
```{r, echo=TRUE}
# this is where your R code goes
head(predictions, 3)
```
### Question 6
I would take the natural logarithm of the data so that we can
convert it to a additive model then difference the model to
remove trend/seasonal components to make the process 
stationary.
