---
title: 'STAT 443: Lab 6'
author: 'Aronn Grant Laurel (21232475)'
date: "14 February, 2022"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#if you do not have the package, type install.packages("name_of_the_package")
library(knitr)
library(tseries)
library(zoo)
```

### Question 1
(a) 

```{r, echo=TRUE}
# this is where your R code goes

# a) Extract data & make into time series object
temp <- read.csv("TempPG.csv")

summer <- temp[, c("Year", "Summer")]

summer_ts <- ts(summer$Summer, 
                start = min(summer$Year), 
                frequency = 1)


```

(b)

```{r, echo=TRUE}
# this is where your R code goes
fig.height=8
par(mfrow = c(3, 1), mar = c(4, 4, 2.5, 1))

# Time series plot
plot(summer_ts, 
     main = "Summer Minimum Temperatures", 
     ylab = "Temperature",
     xlab = "Year", 
     col = "blue")

# sample acf
acf(summer_ts,
    main = "Autocorrelation Function (ACF) of Summer Temperatures",      col = "red")

# sample pacf
pacf(summer_ts,
     main = "Partial Autocorrelation Function (PACF) of Summer Temperatures",
     col = "red")


```
(c)
For our time series plot, we see an noticable increasing trend over 
time from 1920 to 2000 meaning that there is a general increase in
summer minimum temperature over the years.

For the ACF plot, we can observe slowly decaying positive 
autocorrelation at lags 0-15, suggesting a strong trend in the time 
series.

For the PACF plot, we see significant autocorrelation at lag 1, 2,
and 5 while the other lags seem to stay within the boundary line.

(d)
In our ACF function, we see patterns throughout with no cut off in
patterns, therefore q = 0.
In our PACF function, we observe a cut off at lag 2, therefore p =2

Hence, I believe an ideal model would be ARMA(2,0)

### Question 2

(a)
```{r, echo=TRUE}
# this is where your R code goes
fit <- arima(summer_ts, order = c(2, 0, 0))
fit
```

X_t = 7.1615 + 0.4297X_t-1 + 0.3466X_t-2 + Z_t

### Question 3

```{r, echo=TRUE}
# this is where your R code goes
confint(fit)
```
We can observe the true values of ar1, ar2, and the true mean
to be between the values given above.

### Question 4




```{r, echo=TRUE}
# this is where your R code goes
fig.height= 20
# fig.width = 20

par(mfrow = c(3, 1), mar = c(4, 4, 2.5, 1))
par(fin = c(10, 10))

tsdiag(fit,
       gof.lag = 20)



```
We observe fluctuations and osciliations from 2 to -2 for our 
standardized residual plot.

For our ACF plot, we can see an osciliation within the boundary for
all lags starting lag = 1. Mo significant autocorrelation at any lag
suggests that our model captures the temporal dependence in our data well.

For our p values, we see most p-values are well above 0.05 across 
different lags which indicates that we fail to reject the null 
hypothesis of no autocorrelation.

Overall, I believe my ARMA(2,0) model fits well because 
there are no significant autocorrelation in our residuals and it 
does not exhibit any patterns.